Cooperative Kernels: Safe Multitasking for Blocking Algorithms on GPUs

### 2016-11-11 write-ups/PLDI/paper.tex
Use section macros.

Placing the $\offerkill$ and $\offerfork$ primitives requires manual
effort and care, but our experience porting a representative set of
GPU-accelerated irregular algorithms to use cooperative kernels
suggests that this is straightforward in practice.  We discuss this in
detail in Section~\ref{sec:portingalgorithms}.

@@

\item \emph{Cooperative kernels}, an extension to the GPU programming model to support the scheduling requirements of blocking algorithms.  We describe the semantics of the programming model in detail (Section~\ref{sec:cooperativekernels}).

\item A proof-of-concept portable implementation of cooperative
  kernels on top of the OpenCL 2.0 programming model
  (Section~\ref{sec:implementation}).

\item An experience report porting \NumAlgorithms{} irregular GPU
  algorithms to use cooperative kernels, and experimental results
  assessing the performance overhead associated with cooperation and
  the responsiveness to high priority tasks that cooperation enables,
  across four GPUs from two vendors (Section~\ref{sec:experiments}).

@@

We begin by providing an overview of the OpenCL programming model and
two motivating examples of irregular algorithms (Section~\ref{sec:background}).  At the end of the paper we discuss related work (Section~\ref{sec:relatedwork}) and avenues for future work (Section~\ref{sec:conclusion}).

@@

Fine-grained communication between threads, and also with the host in
the case of SVM, is enabled via a set of atomic data types and
operations, based on the C11 memory model~\cite{...}.  \ADComment{Could cut:} In particular,
atomic SVM operations enable fine-grained communication between host
and device.  Our implementation of cooperative kernels
(Section~\ref{sec:implementation}) depends on this recent OpenCL
feature.

@@

Although not depicted here, concurrent accesses to each pool inside
$\mathsf{more\_work}$ and $\mathsf{pop\_or\_steal}$ are guarded by a
mutex, implemented using atomic compare and swap operations on
global memory.  As a result, the kernel presents two opportunities for
spin-waiting: spinning to obtain a mutex, and spinning in the main
kernel loop to obtain a task to process.  The algorithm thus depends
on a fair scheduler, which, as discussed in Section~\ref{sec:intro},
is not provided by current GPU programming models.

@@

In Section~\ref{sec:programmingguidelines} we show how the work stealing
and graph traversal examples of Figures~\ref{fig:workstealing} and~\ref{fig:graphsearch} can be
updated to use our cooperative kernels programming model to resolve
the scheduling issue.

@@

We describe the semantics of cooperative kernels
(Section~\ref{sec:semantics}) and summarise the important
differences between programming with regular vs.\ cooperative kernels
by showing how the motivating examples of \mysec\ref{sec:openclexamples} can be adapted to use cooperative kernels
(Section~\ref{sec:programmingguidelines}).
We then focus on the non-functional
properties that a developer of a cooperative kernel, and the
implementer of a scheduler for cooperative kernels, should strive for
(Section~\ref{sec:nonfunctional}).  The semantics of cooperative
kernels has been guided by the practical applications we have studied
(described in Section~\ref{sec:portingalgorithms}), and we discuss several cases where we
might have taken different and also reasonable semantic decisions (Section~\ref{sec:semanticalternatives}).
We conclude the section by discussing the backwards-compatible nature
of cooperative kernels with respect to blocking algorithms designed to
run on current GPU platforms
(Section~\ref{sec:backwardscompatibility}).

@@

As with a regular OpenCL kernel (see Section~\ref{sec:opencl}), a
cooperative kernel is launched by the host application.  The host
application passes parameters to the kernel and specifies a desired
number of workgroups, each consisting of a specified number of
threads.  The parameters to a cooperative kernel are \emph{immutable}, as if
they were qualified with the \texttt{const} keyword (though pointer
parameters can refer to mutable data).

@@

The $\getgroupid()$ function retrieves the id of a workgroup as usual
(see Section~\ref{sec:opencl}).  When executed by a cooperative
kernel, $\getnumgroups()$ returns $M$, the \emph{current} number of
active workgroups.  This is in contrast to $\getnumgroups()$ for
regular kernels, which returns the fixed number of workgroups that
were requested at kernel launch time.

@@

Similar to the OpenCL workgroup $\mathsf{barrier}$ primitive,
$\offerkill$, is a workgroup-level function---it must be encountered
uniformly by all threads in a workgroup (see
Section~\ref{sec:opencl}).

@@

In this formulation of cooperative kernels, only the active workgroup
with the largest id can be killed, and workgroup 0 can never be
killed.  We discuss the rationale for these choices, as well as
possible alternative choices, in
Section~\ref{sec:semanticalternatives}.

@@

In effect, thread 0 of the forking workgroup transmits the relevant
portion of its local state to the threads of the forked workgroups.
We discuss in Section~\ref{sec:semanticalternatives} the rationale for
having all new threads obtain this local state portion from thread 0,
as well as our reasons for selecting which variables to transmit via
annotations rather than transmitting the entire thread state.

@@

Because $\resizingglobalbarrier$ can be implemented in the above
manner, we do not regard it conceptually as a primitive of the
cooperative kernels programming model.  However, we discuss in
Section~\ref{sec:resizingbarrier} how $\resizingglobalbarrier$ can be
implemented significantly more efficiently if it is treated as a
primitive operation that interacts directly with the scheduler.

@@

The situation is more stable if a cooperative kernel does not directly
issue $\offerkill$ and $\offerfork$ calls directly.  In this case,
only calls to $\resizingglobalbarrier$ can cause the number of active
workgroups to change and, as discussed in Section~\ref{sec:semantics},
this function causes synchronisation between all threads and causes
the number of active workgroups to change atomically.  At any point
during execution, the threads of a kernel are executing between some
pair of resizing barrier calls, which we call a \emph{resizing barrier
  intervals} (here we consider the kernel entry and exit points
conceptually to be special cases of resizing barriers).  The number of
workgroups executing the kernel is thus constant within each resizing
barrier interval, i.e.\ the result returned by repeated calls to
$\getnumgroups()$ will be identical during such an interval; the same holds for $\getglobalsize()$.

@@

The semantics presented in Section~\ref{sec:semantics} describe the envelope of
behaviour that a developer of a cooperative kernel should be prepared
for.
%
However, the aim of cooperative kernels is to find a balance that
allows \emph{efficient execution} of long-running kernels
(e.g.\ implementing irregular data-parallel algorithms), and
\emph{responsive multi-tasking} so that the GPU can be shared between
one or more cooperative kernels and a number of short, high-priority
tasks with soft real-time constraints.

>>>>

Placing the $\offerkill$ and $\offerfork$ primitives requires manual
effort and care, but our experience porting a representative set of
GPU-accelerated irregular algorithms to use cooperative kernels
suggests that this is straightforward in practice.  We discuss this in
detail in \mysec\ref{sec:portingalgorithms}.

@@

\item \emph{Cooperative kernels}, an extension to the GPU programming model to support the scheduling requirements of blocking algorithms.  We describe the semantics of the programming model in detail (\mysec\ref{sec:cooperativekernels}).

\item A proof-of-concept portable implementation of cooperative
  kernels on top of the OpenCL 2.0 programming model
  (\mysec\ref{sec:implementation}).

\item An experience report porting \NumAlgorithms{} irregular GPU
  algorithms to use cooperative kernels, and experimental results
  assessing the performance overhead associated with cooperation and
  the responsiveness to high priority tasks that cooperation enables,
  across four GPUs from two vendors (\mysec\ref{sec:experiments}).

@@

We begin by providing an overview of the OpenCL programming model and
two motivating examples of irregular algorithms (\mysec\ref{sec:background}).  At the end of the paper we discuss related work (\mysec\ref{sec:relatedwork}) and avenues for future work (\mysec\ref{sec:conclusion}).

@@

Fine-grained communication between threads, and also with the host in
the case of SVM, is enabled via a set of atomic data types and
operations, based on the C11 memory model~\cite{...}.  \ADComment{Could cut:} In particular,
atomic SVM operations enable fine-grained communication between host
and device.  Our implementation of cooperative kernels
(\mysec\ref{sec:implementation}) depends on this recent OpenCL
feature.

@@

Although not depicted here, concurrent accesses to each pool inside
$\mathsf{more\_work}$ and $\mathsf{pop\_or\_steal}$ are guarded by a
mutex, implemented using atomic compare and swap operations on
global memory.  As a result, the kernel presents two opportunities for
spin-waiting: spinning to obtain a mutex, and spinning in the main
kernel loop to obtain a task to process.  The algorithm thus depends
on a fair scheduler, which, as discussed in \mysec\ref{sec:intro},
is not provided by current GPU programming models.

@@

In \mysec\ref{sec:programmingguidelines} we show how the work stealing
and graph traversal examples of Figures~\ref{fig:workstealing} and~\ref{fig:graphsearch} can be
updated to use our cooperative kernels programming model to resolve
the scheduling issue.

@@

We describe the semantics of cooperative kernels
(\mysec\ref{sec:semantics}) and summarise the important
differences between programming with regular vs.\ cooperative kernels
by showing how the motivating examples of \mysec\ref{sec:openclexamples} can be adapted to use cooperative kernels
(\mysec\ref{sec:programmingguidelines}).
We then focus on the non-functional
properties that a developer of a cooperative kernel, and the
implementer of a scheduler for cooperative kernels, should strive for
(\mysec\ref{sec:nonfunctional}).  The semantics of cooperative
kernels has been guided by the practical applications we have studied
(described in \mysec\ref{sec:portingalgorithms}), and we discuss several cases where we
might have taken different and also reasonable semantic decisions (\mysec\ref{sec:semanticalternatives}).
We conclude the section by discussing the backwards-compatible nature
of cooperative kernels with respect to blocking algorithms designed to
run on current GPU platforms
(\mysec\ref{sec:backwardscompatibility}).

@@

As with a regular OpenCL kernel (see \mysec\ref{sec:opencl}), a
cooperative kernel is launched by the host application.  The host
application passes parameters to the kernel and specifies a desired
number of workgroups, each consisting of a specified number of
threads.  The parameters to a cooperative kernel are \emph{immutable}, as if
they were qualified with the \texttt{const} keyword (though pointer
parameters can refer to mutable data).

@@

The $\getgroupid()$ function retrieves the id of a workgroup as usual
(see \mysec\ref{sec:opencl}).  When executed by a cooperative
kernel, $\getnumgroups()$ returns $M$, the \emph{current} number of
active workgroups.  This is in contrast to $\getnumgroups()$ for
regular kernels, which returns the fixed number of workgroups that
were requested at kernel launch time.

@@

Similar to the OpenCL workgroup $\mathsf{barrier}$ primitive,
$\offerkill$, is a workgroup-level function---it must be encountered
uniformly by all threads in a workgroup (see
\mysec\ref{sec:opencl}).

@@

In this formulation of cooperative kernels, only the active workgroup
with the largest id can be killed, and workgroup 0 can never be
killed.  We discuss the rationale for these choices, as well as
possible alternative choices, in
\mysec\ref{sec:semanticalternatives}.

@@

In effect, thread 0 of the forking workgroup transmits the relevant
portion of its local state to the threads of the forked workgroups.
We discuss in \mysec\ref{sec:semanticalternatives} the rationale for
having all new threads obtain this local state portion from thread 0,
as well as our reasons for selecting which variables to transmit via
annotations rather than transmitting the entire thread state.

@@

Because $\resizingglobalbarrier$ can be implemented in the above
manner, we do not regard it conceptually as a primitive of the
cooperative kernels programming model.  However, we discuss in
\mysec\ref{sec:resizingbarrier} how $\resizingglobalbarrier$ can be
implemented significantly more efficiently if it is treated as a
primitive operation that interacts directly with the scheduler.

@@

The situation is more stable if a cooperative kernel does not directly
issue $\offerkill$ and $\offerfork$ calls directly.  In this case,
only calls to $\resizingglobalbarrier$ can cause the number of active
workgroups to change and, as discussed in \mysec\ref{sec:semantics},
this function causes synchronisation between all threads and causes
the number of active workgroups to change atomically.  At any point
during execution, the threads of a kernel are executing between some
pair of resizing barrier calls, which we call a \emph{resizing barrier
  intervals} (here we consider the kernel entry and exit points
conceptually to be special cases of resizing barriers).  The number of
workgroups executing the kernel is thus constant within each resizing
barrier interval, i.e.\ the result returned by repeated calls to
$\getnumgroups()$ will be identical during such an interval; the same holds for $\getglobalsize()$.

@@

The semantics presented in \mysec\ref{sec:semantics} describe the envelope of
behaviour that a developer of a cooperative kernel should be prepared
for.
%
However, the aim of cooperative kernels is to find a balance that
allows \emph{efficient execution} of long-running kernels
(e.g.\ implementing irregular data-parallel algorithms), and
\emph{responsive multi-tasking} so that the GPU can be shared between
one or more cooperative kernels and a number of short, high-priority
tasks with soft real-time constraints.
